---
title: "Statistical Tests"
subtitle: "Computer Lab 3"
abstract: |
  The purpose of today's lab is for you to:

  - practice the basic concepts of hypothesis testing
  (e.g., significance level and power) and what conclusions can
  be drawn from the analysis
  - familiarize yourself with some of the functions available in R regarding various basic statistical tests
  - work on the course's Mini Project II.
author: 
  - name: "Your Name"
date: today
toc: true
number-sections: true
highlight-style: pygments
format: 
  html:
    theme: cosmo
    code-fold: false
  pdf: 
    geometry: 
      - top=30mm
      - left=30mm
  docx: default
---

## Preparation Tasks {.unnumbered}

> Review the concepts of *hypotheses, significance level, power function,* and *model with matched data and model with two independent samples*.
>
> You should have completed the following tasks *before* you come to the lab.

:::{.callout-warning icon=false}
### Homework 1
To practice the basic concepts in hypothesis testing, complete exercises **Dig:4.4.1(error rates)\_3, \_4, and \_5 in exercise 7**.
:::

:::{.callout-warning icon=false}
### Homework 2
To practice different models, do **Dig:5.2\_1 and \_2**.
:::

:::{.callout-warning icon=false}
### Before You Start
Load R functions needed for the lab. These functions come from the R package **Räkna med variation**, but to avoid installation problems, we have extracted the ones you will use and saved them in a file. You "inform" R that these functions exist by running the following code:

```{r}
#| message: false
source("kod/funktioner_raknamedvariation_light.R")
```
:::
# Basic Concepts in Hypothesis Testing

> Medication can reduce salivary gland function, which is a risk factor for cavities and other oral diseases. Saliva stimulated by chewing was measured for 5 minutes in 7 randomly selected patients, all receiving the same medication. The normal saliva amount under these conditions is 1 ml/min, and dry mouth is considered when the saliva amount is below 0.7 ml/min. The model assumed that the saliva amount is normally distributed with mean $\mu$ and standard deviation $\sigma$, where $\sigma$ is considered to be 0.5 ml/min. Interesting questions include:
> 
> - Do the data support our suspicion that the medication lowers saliva production?
> - If the medication results in an average saliva production of 0.8 ml/min, how likely are we to miss the reduced saliva production with our test?
> - How many patients should we measure if we want the test to detect reduced saliva production at 0.7 ml/min with a probability of 0.95? 
>
> On the course website, you will find data in the file `salivary.RData`. Concise answers to the questions posed in the tasks are provided at the end of this part of the guide.

:::{.callout-note icon=false}
### Task 1.1
First, we want to investigate if the data from the 7 patients support our suspicion that the medication lowers saliva production. Set up a model and appropriate hypotheses.
:::

**Model**: <!-- Remove this and specify the model here  -->

**Hypotheses**: <!-- Remove this and specify the hypotheses here  -->

:::{.callout-note icon=false}
### Task 1.2
Calculate the mean of the measurements.
:::

```{r}
# Write your R code here
```

**Answer**: <!-- Remove this and write down the mean of the sample -->

:::{.callout-note icon=false}
### Task 1.3
Use the routine `hypotes` to illustrate the test's critical region when the test is performed at a significance level of $\alpha = 0.05$. The relevant command is `hypotes(`$\sigma$, $n$, $\mu_0$, $\alpha$, `H1-direction)`. With hypotheses $H_0: \mu = 1$ and $H_1: \mu < 1$, the command becomes `hypotes(0.5, 7, 1, 0.05, '<')`. Use the routine with the hypotheses you have formulated. (Ignore any error messages that may occur and look at the figure.)
:::

```{r}
# Write your R code here
```


:::{.callout-note icon=false}
### Task 1.4
The routine marks the critical region and provides a value $k$
which is the boundary of the region. How has $k$ been calculated? Set up the mathematical formula for $k$ and use it to calculate $k$ yourself.
:::

**Equation for $k$**: <!-- Remove this and write the expression -->

**Own calculation of $k$**:
```{r}
# Write your R code here
```


:::{.callout-note icon=false}
### Task 1.5

Use your calculated sample mean to conduct the test. What is your conclusion about the null hypothesis, $H_0$?
:::

**Answer:** <!-- remove this and write your conclusions in plain text -->

:::{.callout-note icon=false}
### Task 1.6
What is your concrete interpretation of the significance level $\alpha = 0.05$ in this example?
:::

**Answer:** <!-- remove this and write your conclusions in plain text -->

:::{.callout-note icon=false}
### Task 1.7
Examine how the critical region changes when you change the significance level to $\alpha = 0.01$. What is your conclusion now?
:::

```{r}
# write your R code here
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

# Test Power and Power Function

> Now assume that the average saliva secretion in the risk group is 0.8. Then, of course, $H_0: \mu = 1$ is false, and we want our test to detect this and reject this hypothesis in favor of the hypothesis $H_1: \mu < 1$. The probability that the test actually accomplishes this is called the test power at the point 0.8. Use the routine `hypotes` to illustrate the test power at the point 0.8. The command now is `hypotes(`$\sigma$, $n$, $\mu_0$, $\alpha$, H1-direction, true $\mu$`), so in this case, you write `hypotes(0.5, 7, 1, 0.05, '<', 0.8)`.

```{r}
#| fig-width: 7
#| fig-height: 7
#| out-width: 80%
hypotes(0.5, 7, 1, 0.05, '<', 0.8)
```

:::{.callout-note icon=false}

### Task 2.1
The routine gives you another figure which, besides the significance level $\alpha$ (type I error risk), also shows $\beta$ (type II error risk). What is the concrete interpretation of $\beta$ in this example? How does $\beta$ relate to the test power?

Generally, the test power at the point $\mu$ is the probability that the null hypothesis is rejected given that $\mu$ is the true average saliva secretion in the risk group, i.e.,

$$
S(\mu)=P(\text{reject } H_0 | \mu)
$$

Note that the power depends on the value of $\mu$. In this example, the smaller $\mu$ is relative to $\mu_0= 1$, the greater the chance that the test will detect that $H_0$ is not true. Therefore, it is interesting to study the power as a function of $\mu$, and this function is often denoted $S(\mu)$.

The routine \texttt{styrkefkn} draws the power function. The command is \texttt{styrkefkn($\sigma$, $n$,   $\mu_0$, $\alpha$, H1-direction, true $\mu$)} if you want to draw the function and mark a specific $\mu$ value. So, you use the command `styrka(0.5, 7, 1, 0.05, '<', 0.8)`, which gives you the previous figure plus the power as a function of $\mu$.
:::

```{r}
#| fig-width: 7
#| fig-height: 7
#| out-width: 80%
styrka(0.5, 7, 1, 0.05, '<', 0.8)
```

<!-- remove this and write your conclusions in plain text -->

:::{.callout-note icon=false}
### Task 2.2
Based on the power function, estimate the probability that with our test, we will detect that a group that should be classified as having dry mouth ($\mu= 0.7$) has reduced saliva production.
:::

```{r}
# write your R code here
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

:::{.callout-note icon=false}
### Task 2.3
How many patients should we measure if we want to have a 0.95 probability of detecting that individuals with dry mouth have reduced saliva production? Hint: Try different values of $n$ in `styrkefkn`.
:::

```{r}
# write your R code here
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

## Answer to the example with dry mouth {.unnumbered}

 1. Part 1  
  1.1 $H_0: \mu = 1$;  $H_1: \mu < 1$  
  1.2 $k=\mu_0 - z_{1-\alpha}\, \frac{\sigma}{\sqrt{n}} =  1-1.6445 \cdot \frac{0.05}{\sqrt{7}}=0.689$  
  1.5 Since the mean is $< 0.689$, $H_0$ is rejected at the 0.05 level  
  1.6 There is a 5% chance that we claim a person in the risk group has reduced saliva production when it is actually normal
  1.7 $H_0$ cannot be rejected at the 0.01 level
 2. Part 2  
  2.1 $\beta = P(\text{not reject } H_0|\mu)$, i.e., $\beta$ when the true expected value is 0.8 is 1 minus the power at the point 0.8 = 1- S(0.8)  
  2.2 The power at the point 0.7, S(0.7), can be estimated as 0.48 according to the figure  
  2.3 $n=30$ patients are required for the power to be 0.95 at the point 0.7  



## Some Statistical Tests with Samples {.unnumbered}

> Concise answers to the questions posed in the exercises are found at the end of this section of the guide. The data materials used are albumin, leaf chlorophyll, and dust, which you can find on Canvas.

# Test of Population Mean (t-test)

> A mixture of blood serum contains exactly 42 g of albumin per liter. Two laboratories (A and B) make six determinations each of the concentration. We want to investigate if there is any systematic deviation from the true value (42 g/l) in each of these two groups.

```{r}
load("data/lab3_mini2_filer/albumin.RData")
```

:::{.callout-note icon=false}
### Exercise 3.1
Specify the model for A and B and specify hypotheses to test if there is a systematic deviation from the true value in each group. Hint: it should be a two-sided alternative hypothesis.
:::

**Model for group A:** Let $X$ be the concentration of g of albumin per liter in measurements from laboratory A. The mean and variance for $X$ are $\mu_x$ and $\sigma^2_x$. Assume that measurements are independent and identically distributed. The mean is estimated by the sample mean $\hat{\mu}_x=\bar{x}$. According to the central limit theorem, $\bar{x} \overset{A}{\sim} N(\mu_x,\frac{\sigma_x}{\sqrt{n_x}})$.

**Hypotheses for group A:** <!-- remove this and write your hypotheses-->

**Model for group B:** Let $Y$ be the concentration of g of albumin per liter in measurements from laboratory B. <!-- remove this and continue writing your model, specify assumptions-->

**Hypotheses for group B:** <!-- remove this and write your hypotheses-->

:::{.callout-note icon=false}
### Exercise 3.2
In R, perform these two t-tests using the following routine `t-test`. Interpret the outputs when you run the command. What are the p-values and what are the conclusions for each group? What are the confidence intervals for the expected concentrations?
:::

```{r}
t.test(AlbuminA, mu=42)
t.test(AlbuminB, mu=42)
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

:::{.callout-note icon=false}
### Exercise 3.3
Suppose we want to investigate if **laboratory B** underestimates or overestimates concentrations. In R, this is done as a one-sided test using the routine `t-test`.

Interpret the output for the one-sided test for laboratory B. What happens to the p-values and intervals when moving from one-sided to two-sided alternatives?
:::

```{r}
t.test(AlbuminB, mu=42, alternative="less")
t.test(AlbuminB, mu=42, alternative="greater")
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

# Comparison of Means in Two Populations (t-test for Two Independent Samples)

> Algae were allowed to grow under light and dark conditions, and then the chlorophyll content was measured.

```{r}
load("data/lab3_mini2_filer/alger.RData")
```

:::{.callout-note icon=false}
### Exercise 4.1
Make a graphical description of the differences. What does the boxplot show?
:::

```{r}
boxplot(Alger$Klorofyll ~ Alger$Grupp)
```

**Answer:** <!-- remove this and write your conclusions in plain text -->


:::{.callout-note icon=false}
### Exercise 4.2
Set up the model and hypotheses, and investigate with a t-test if there are differences in the expected chlorophyll content between the two groups.

This can be done by adding the group variable in the `t.test` routine. Interpret the output.
:::

**Model:** Let $X$ be the growth of algae in dark conditions with mean $\mu_x$ and variance $\sigma^2_x$, and $Y$ be the growth of algae in light conditions with mean $\mu_y$ and variance $\sigma^2_y$. The means are estimated by sample means $\hat{\mu_x}=\bar{x}$ and $\hat{\mu_y}=\bar{y}$. According to the central limit theorem, the estimation of each mean will be approximately normally distributed, i.e., $\bar{x} \overset{A}{\sim} N(\mu_x,\frac{\sigma_x}{\sqrt{n_x}})$ and $\bar{y} \overset{A}{\sim} N(\mu_y,\frac{\sigma_y}{\sqrt{n_y}})$.

**Hypotheses:** $H_0: \mu_x = \mu_y$ vs. $H_1: \mu_x \neq \mu_y$ 

```{r}
t.test(Alger$Klorofyll ~ Alger$Grupp)
```

**Answer:** <!-- remove this and write your conclusions in plain text -->


:::{.callout-note icon=false}
### Exercise 4.3
If we don't specify otherwise, the `t.test` routine assumes that the variances in the two groups, i.e., $\sigma^2_x$ and $\sigma^2_y$, are **unequal** and adjusts for it. If we know (or assume) that the variances are equal, we can utilize that to obtain higher degrees of freedom since there's one less parameter to estimate.

Investigate if the variances are equal using a so-called $\chi^2$-test. This is done in R using the `var.test` routine. Interpret the output!
:::

**Hypotheses**: $H_0: \sigma^2_x = \sigma^2_y$ vs. $H_1: \sigma^2_x \neq \sigma^2_y$
```{r}
var.test(Alger$Klorofyll ~ Alger$Grupp)
```

**Answer:** <!-- remove this and write your conclusions in plain text -->


:::{.callout-note icon=false}
### Exercise 4.4
Repeat the t-test assuming equal variances by adding `var.equal=TRUE` in the R routine. What are your conclusions now?
:::

**Assumption**: $\sigma^2=\sigma^2_x = \sigma^2_y$

```{r}
t.test(Alger$Klorofyll ~ Alger$Grupp, var.equal=TRUE)
```

**Answer:** <!-- remove this and write your conclusions in plain text -->

# Test with Matched Data (Paired Samples)

> Measurements of nitrogen load have been taken during spring and summer in a number of ponds. If one wants to compare the nitrogen load between the two seasons, a reasonable model is "paired samples."

```{r}
load("data/lab3_mini2_filer/dammar.RData")
```

```{r}
boxplot(Dammar$N_belast_V, Dammar$N_belast_S)
```


```{r}
boxplot(Dammar$N_belast_V-Dammar$N_belast_S)
abline(a=0,b=0,col='blue')
```

**Model:** Let $D$ be the difference in measurements between spring and summer for a pond. The random variable $D$ has mean $\mu$ and variance $\sigma^2$. We create a sample $(d_1,...,d_n)$ for differences in measurement values. The mean is estimated by the sample mean, $\hat{\mu}=\bar{d}$. According to the central limit theorem, $\bar{d} \overset{A}{\sim} N(\mu,\frac{\sigma}{\sqrt{n}})$. 


:::{.callout-note icon=false}
### Exercise 5.1
Is there a difference in means between spring and summer? 

We test it with the following **hypotheses:** $H_0: \mu = 0$ vs. $H_1: \mu \neq 0$

This is a common test, and one does not need to calculate a new sample. To inform `t.test` that the data is matched in pairs, one adds `paired=TRUE` as an argument.
:::


```{r}
t.test(Dammar$N_belast_V, Dammar$N_belast_S, paired=TRUE)
```

**Answer**: <!-- remove this and write your conclusions in plain text -->


## Answers to Some of the Questions in This Section: {.unnumbered}  

1. Answer  
  3.2 A: p-value$=0.033$; B: p-value$=0.081$; We can detect a difference for A but not for B. 95% confidence intervals: A: $(42.06,\,42.94)$; B: $(35.69,\,42.51)$  
  4.3 The variances are not different (p-value$=0.48$)  
  4.4 t-test: p-value$=0.009$; We can detect a difference in chlorophyll.  
  5.1 p-value$=0.262$ We cannot detect any difference.  